The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-14 manuscript/manuscript.Rnw
cite GH1 gene paper

The \ac{GH} family has expanded in anthropoid primates, resulting in
approximately 5 genes in \human (\textit{GH1}, \textit{GH2},
\textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at least three in
\bonobo (two \textit{GH2} and \textit{GH1}; owing to the similarity
between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
reliably distinguish transcripts between the two). All but
\textit{GH1} show substantial expression in the \human placenta
(\cref{fig:csh1_evolution_tree}). There is some evidence that
increased expression of these genes may be correlated with overall
birth weight of offspring in \human
\cite{Männik.ea2010:Differentialexpressionprofileof}, and it is
possible that the extreme expression levels of these genes may be due
to high birth weight. \human placental \ac{GH} genes are involved in
increasing blood flow, replacing maternal pituitary expression of
\ac{GH}, as well as trophoblast invasion
\cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth},
and it is likely that they have the same function in \bonobo.

>>>>

The \ac{GH} family has expanded in anthropoid primates, resulting in
approximately 5 genes in \human (\textit{GH1}, \textit{GH2},
\textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at least three in
\bonobo (two \textit{GH2} and \textit{GH1}; owing to the similarity
between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
reliably distinguish transcripts between the two). All show
substantial expression in the \human placenta
(\cref{fig:csh1_evolution_tree}). The expression of the \ac{GH} family
in placenta appears to be primate specific, and likely existed at the
time of the last common ancestor between all extant anthropoids
\cite{Papper.ea2009:Ancientoriginofplacental}. There is some evidence
that increased expression of these genes may be correlated with
overall birth weight of offspring in \human
\cite{Männik.ea2010:Differentialexpressionprofileof}, and it is
possible that the extreme expression levels of these genes may be due
to high birth weight. \human placental \ac{GH} genes are involved in
increasing blood flow, replacing maternal pituitary expression of
\ac{GH}, as well as trophoblast invasion
\cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth},
and it is likely that they have the same function in \bonobo.
